\morus: Biases in the Stream

### 2018-05-10 morusAC_01_Introduction.tex
homogenized references

\subsubsection*{Paper Outline.}
This paper is organized as follows.
We first provide a brief description of \morus in \autoref{sec/Preliminaries}.
In \autoref{sec/introminimorus}, we introduce \minimorus, an abstraction of \morus based on a certain class of rotational symmetries.
We analyze this simplified scheme in \autoref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \autoref{sec/fulltrails}, showing a bias in the keystream over 5 rounds,
and discuss the implications of our observation for the security of \morus in \autoref{sec/Discussion}.
In \autoref{sec/IniFin}, we present our results on the security of \morus with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \autoref{sec/Conclusion}.

>>>>

\subsubsection*{Article Outline.}
This paper is organized as follows.
We first provide a brief description of \morus in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \minimorus, an abstraction of \morus based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a bias in the keystream over 5 rounds,
and discuss the implications of our observation for the security of \morus in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \morus with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

### 2018-05-10 morusAC_02_Preliminaries.tex
minor update
homogenized references

\StateUpdate{} is composed of 5 rounds with similar operations. The additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round uses the word-wise left circular shift operation, which divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a left circular shift operation on each $w$-bit word. The word-wise circular shift constants $b_i$ in round $i$ are defined in Table~\ref{Tbl:rcon}. Left circular shifts on a whole $q$-bit register are also used. The register-wise circular shift constants are also listed in Table~\ref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Left Circular Shift Constants of MORUS} \label{Tbl:rcon}
\begin{tabular}{r||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{Word-wise circular shift} & \multicolumn{5}{c}{Register-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\cipher{MORUS640} &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\cipher{MORUS1280}& 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}

>>>>

\StateUpdate{} is composed of 5 rounds with similar operations. The additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round uses the word-wise left circular shift operation, which divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a left circular shift operation on each $w$-bit word. The word-wise circular shift constants $b_i$ in round $i$ are defined in \Cref{Tbl:rcon}. Left circular shifts on a whole $q$-bit register are also used. The register-wise circular shift constants are also listed in \Cref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Left Circular Shift Constants of MORUS} \label{Tbl:rcon}
\begin{tabular}{r||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{Word-wise circular shift} & \multicolumn{5}{c}{Register-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\cipher{MORUS640} &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\cipher{MORUS1280}& 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}

### 2018-05-10 morusAC_07_IniFin.tex
homogenized references

\subsubsection{Differential Trails.}
Recall that the message input during the finalization of \cipher{MORUS1280} is $adlen\|msglen\|0^{128}$ where $adlen$ and $msglen$ are 64-bit strings. We set $\Delta adlen$ to be low Hamming weight, e.g.~{\tt 0x0000000000000001}. 
%Hereafter, we use {\tt typefont} denote hexadecimal numbers.
This difference propagates during 3 steps as specified in Table~\ref{Tbl:fin_diff}.

Recall that each step consists of 5 rounds and the input message is absorbed to the state in rounds 2 to 5. The trail in Table~\ref{Tbl:fin_diff} initially does not have any difference and the same continues even after round 1. Differences start to appear from round 2 and they will go through the bitwise-AND operation from round 4. We need to pay 1-bit to control each active AND gate. The probability evaluation for round 15 can be ignored because in this round only $S_4$ is non-linearly updated, while $S_4$ is never used for computing the tag. Finally, bitwise-AND in the tag computation is taken into account. Note that the tag is only 128 LSBs, thus the number of active AND gates should be counted only in those bits. As shown in Table~\ref{Tbl:fin_diff}, we can have a particular tag difference $\Delta T$ with probability $2^{-88}$. Thus after observing $A$ and corresponding $T$, $A\|0$ and $(T \oplus \Delta T)$ is a valid pair with probability $2^{-88}$.

@@

\subsubsection{Remarks.} The fact that $S_4$ updated in the last round is not used to compute a tag implies that the MORUS finalization generally includes unnecessary computations with respect to security. It may be interesting to tweak the design such that the tag can also depend on $S_4$. Indeed in Table~\ref{Tbl:fin_diff}, we can observe some jump-up of the probability in the tag computation. This is because the non-linearly involved terms are $S_2 \cdot S_3$, and $S_3$ that was updated 2-round before has a high Hamming weight. In this sense, involving $S_4$ in non-linear terms of the tag computation imposes more difficulties for the attacker.

@@

\subsubsection{Full Diffusion Rounds.}
We found that {\tt StepUpdate} was designed to have good diffusion in the forward direction. Thus, once the state is recovered, the attacker can perform the partial computation in the backward direction longer than the forward direction. We set $G_0$ and $G_1$ as follows.
\begin{align*}
G_0 &= \{1, 2, \cdots, 127\} & \textrm {Bit position 0 is unknown.}\\
G_1 &= \{0, 1, \cdots, 7, 9, 10, \cdots, 127\} & \textrm {Bit position 8 is unknown.}
\end{align*}
Those will lead to 4 matching bits after the 4-step forward computation and the 6-step backward computation. The analysis of the diffusion is given in Table~\ref{Tbl:ini_diffuse}. In the end, $K$ can be recovered faster than the exhaustive search by 1 bit, i.e.~with complexity $2^{127}$.

@@

As can be seen in Table~\ref{Tbl:ini_diffuse}, the updated register in step $i$ is independent of the update function in step $i+1$ in the forward direction, and starts to impact from step $i+2$. By modifying this point, the diffusion speed can increase faster, which makes this attack harder.

>>>>

\subsubsection{Differential Trails.}
Recall that the message input during the finalization of \cipher{MORUS1280} is $adlen\|msglen\|0^{128}$ where $adlen$ and $msglen$ are 64-bit strings. We set $\Delta adlen$ to be low Hamming weight, e.g.~{\tt 0x0000000000000001}. 
%Hereafter, we use {\tt typefont} denote hexadecimal numbers.
This difference propagates during 3 steps as specified in \Cref{Tbl:fin_diff}.

Recall that each step consists of 5 rounds and the input message is absorbed to the state in rounds 2 to 5. The trail in \Cref{Tbl:fin_diff} initially does not have any difference and the same continues even after round 1. Differences start to appear from round 2 and they will go through the bitwise-AND operation from round 4. We need to pay 1-bit to control each active AND gate. The probability evaluation for round 15 can be ignored because in this round only $S_4$ is non-linearly updated, while $S_4$ is never used for computing the tag. Finally, bitwise-AND in the tag computation is taken into account. Note that the tag is only 128 LSBs, thus the number of active AND gates should be counted only in those bits. As shown in \Cref{Tbl:fin_diff}, we can have a particular tag difference $\Delta T$ with probability $2^{-88}$. Thus after observing $A$ and corresponding $T$, $A\|0$ and $(T \oplus \Delta T)$ is a valid pair with probability $2^{-88}$.

@@

\subsubsection{Remarks.} The fact that $S_4$ updated in the last round is not used to compute a tag implies that the MORUS finalization generally includes unnecessary computations with respect to security. It may be interesting to tweak the design such that the tag can also depend on $S_4$. Indeed in \Cref{Tbl:fin_diff}, we can observe some jump-up of the probability in the tag computation. This is because the non-linearly involved terms are $S_2 \cdot S_3$, and $S_3$ that was updated 2-round before has a high Hamming weight. In this sense, involving $S_4$ in non-linear terms of the tag computation imposes more difficulties for the attacker.

@@

\subsubsection{Full Diffusion Rounds.}
We found that {\tt StepUpdate} was designed to have good diffusion in the forward direction. Thus, once the state is recovered, the attacker can perform the partial computation in the backward direction longer than the forward direction. We set $G_0$ and $G_1$ as follows.
\begin{align*}
G_0 &= \{1, 2, \cdots, 127\} & \textrm {Bit position 0 is unknown.}\\
G_1 &= \{0, 1, \cdots, 7, 9, 10, \cdots, 127\} & \textrm {Bit position 8 is unknown.}
\end{align*}
Those will lead to 4 matching bits after the 4-step forward computation and the 6-step backward computation. The analysis of the diffusion is given in \Cref{Tbl:ini_diffuse}. In the end, $K$ can be recovered faster than the exhaustive search by 1 bit, i.e.~with complexity $2^{127}$.

@@

As can be seen in \Cref{Tbl:ini_diffuse}, the updated register in step $i$ is independent of the update function in step $i+1$ in the forward direction, and starts to impact from step $i+2$. By modifying this point, the diffusion speed can increase faster, which makes this attack harder.

### 2018-05-10 morusAC_08_Conclusion.tex
homogenized references

We have shared an earlier version of this paper with the authors of
\morus and they agree with the technical details of the keystream
bias.  However they consider that it is not a significant weakness in
practice because it requires more than $2^{64}$ ciphertexts bits.  In
the context of the CAESAR competition, we believe that certificational
attacks such as this one should be taken into account, in order to
select a portfolio of candidates that reflect the state of the art in
terms of cryptographic design.

>>>>

We have shared an earlier version of this paper with the authors of
\morus and they agree with the technical details of the keystream
bias.  However they consider that it is not a significant weakness in
practice because it requires more than $2^{64}$ ciphertexts bits.  In
the context of the CAESAR competition, we believe that certificational
attacks such as this one should be taken into account, in order to
select a portfolio of candidates that reflects the state of the art in
terms of cryptographic design.
